Spurious Sentience:\\ Dovelating Inflated AI World Model Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}

### 2024-01-24 main.tex
Updates from Overleaf

Hypotheses in computer science fields related to AI are often implicit. Often framed as informally that a given system is more accurate or efficient compared to other systems, and not typically articulated as they are in other fields, with specific conditions and in some instances numerical values or effect sizes assigned to each competing hypothesis. However with regards to claims that outputs of a system hint towards emerging qualities like AGI, hypotheses necessarily become more explicit: either one interprets outputs as hinting towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).

Confirmation bias in hypothesis testing remains relevant and may manifest as a number of behaviors, as \cite{nickerson1998confirmation} reviews. Scientists may pay little to no attention to competing hypotheses or explanations, e.g. only considering the likelihood that outputs of system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false. Similarly, bias may show when failing to sufficiently articulate a strong enough competing hypothesis, leading to a 'weak' or 'non-risky' test, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}: if the threshold to accept a favored hypothesis is too low, observations consistent with the hypothesis are almost guaranteed, and in turn fail to severely test the claim in question. Thus, one is far more likely to show evidence in favor of their beliefs by posing a weak competing hypothesis.

Related to the formulation of hypotheses is the interpretation of evidence in favor of competing hypotheses, wherein people will interpret identical evidence differently based on their beliefs.

@@

\section{Patterns in Latent Spaces and How to Find Them}

To demonstrate the claim that observing patterns in latent spaces should not generally surprise us, we will now go through a couple of simple examples.

\subsection{Example: Principal Component Analysis}

To understand this better, let us go on a quick detour into economics and look at actual yield curves observed in the US during the Global Financial Crisis (GFC). Figure 1 (a) shows the yield curve of US Treasury bonds on 27 February 2007, which according to CNN was a “brutal day on Wall Street”.2 This followed reports on the previous day of former Federal Reserve Chairman Alan Greenspan’s warning that the US economy was at risk of a recession. The yield curve was inverted with a sharp negative spread between the 10-year and 3-month yields, indicative of the market’s expectation of a recession.

Figure 1 (b) shows the corresponding yield curve during the aftermath of the GFC on 20 April 2009. On that day the influential Time Magazine reported that the “Banking Crisis is Over”. The yield curve was steeply sloped with a positive spread between the 10-year and 3-month yields, indicative of the market’s expectation of a recovery. The overall level of the yield curve was still very low though, indicative of the fact that US economy had not fully recovered at that point.

@@

To compute the principal components we can decompose the matrix of yields 
 into a product of its singular vectors and values: 
. I will not go into the details here, because Professor Gilbert Strang has already done a much better job than I ever could in his Linear Algebra lectures. To put this into the broader context of the article, however, let us simply refer to 
, 
 and 
 as latent embeddings of the yield curve (they are latent because they are not directly observable).

The top panel in Figure 2 shows the first two principal components of the yield curves of US Treasury bonds over time. Vertical stalks indicate the key dates during the onset and aftermath of the crisis, which we discussed above. For both components, we can observe some marked shifts between the two dates - but can we attribute any meaning to these shifts? It turns out we can: for comparison, the bottom panel in Figure 2 shows the average level and spread of the yield curves over time. The first principal component is strongly correlated with the level of the yield curve, while the second principal component is strongly correlated with the spread of the yield curve. To put it in AI-lingo:

Not convinced? Let us use 
 in true autoencoder fashion to reconstruct yield curves from principal components. Let 
 denote the first principal component and consider the following: we keep all other 
 principal components fixed at zero where 
 denotes the total number of maturities; next we traverse the latent space by varying the value of 
 over a fixed grid of length 
 each time storing the full vector 
; finally, we vertically concatenate the vectors and end up with a matrix 
 of dimension 
. To reconstruct yields, we simply multiply 
 by the singular values and right singular vectors: 
.

Figure 3 shows the result of this exercise in the left panel. As we can see, our generated yield curves shift vertically as we traverse the latent space. The right panel of Figure 3 shows the result of a similar exercise, but this time we keep the first principal component fixed at zero and vary the second principal component. This time the slope of our generated yield curves shifts as we traverse the latent space.

\subsection{Example: Deep Learning}

@@

\subsubsection{Data}

To estimate economic growth we will rely on a quarterly series of the real gross domestic product (GDP) provided by the Federal Reserve Bank of St. Louis. The data arrives in terms of levels of real GDP. In order to estimate growth, we will transform the data into log differences. Since our yield curve data is daily, we will need to aggregate it to the quarterly frequency. To do this, we will simply take the average of the daily yields for each maturity. We will also standardize yields since deep learning models tend to perform better with standardized data. Since COVID-19 was a huge structural break, we will also filter out all observations after 2018. Figure 4 shows the pre-processed data.

\subsubsection{Model}

Let 
 denote growth and 
 denote the yield curve at time 
. Then we are interested in a model for 
 conditional on 
. Let 
 denote our model parameters then formally we are interested in maximizing the likelihood 
. To do this, we will use a simple autoencoder architecture that is illustrated in Figure 5. The encoder will consist of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function. The bottleneck layer connecting the encoder to the decoder is a fully connected layer with 6 neurons. This is the compressed latent representation of the yield curve mentioned above. The decoder will consist of two fully connected layers each with a hyperbolic tangent activation function: the first layer will consist of 32 neurons and the second layer will have the same dimension as the input data. The output layer will consist of a single neuron with a linear activation function. The model will be trained using the mean squared error loss function and the Adam optimizer.

\subsubsection{Linear Probe}

The results are shown in Figure 6. The top panel shows the actual GDP growth and fitted values from the autoencoder model. We observe that the model captures the relationship between economic growth and the yield curve reasonably well. As discussed above, we also know that the relationship between economic growth and the yield curve is characterized by two main factors: the level and the spread. Since the model itself is fully characterized by its parameters, we would expect that these two important factors are reflected somewhere in the latent parameter space.

The bottleneck layer seems like a good place to start looking. To get the latent embeddings 
 at time 
, we simply pass the yield curve data 
 through the encoder. Next, we follow [TEGMARK PAPER REFERENCE] and use a linear probe to regress the observed yield curve factors on the latent embeddings. Let 
 denote the vector containing the two factors of interest in time 
: 
 and 
. Formally, we are interested in the following regression model: 
 where 
 denotes the regression parameters. Following [TEGMARK PAPER REFERENCE] we use Ridge regression with 
 set to 
. Using the estimated regression parameters 
 we can then predict the yield curve factors from the latent embeddings: 
.

The results of this experiment are shown in the bottom panel of Figure 6. Solid lines show the observed yield curve factors over time, while dashed lines show predicted values. We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread. To put this in true hype-lingo:

Not just a parrot! Our autoencoder neural network has an implicit understanding of the yield curve factors and their relationship with economic growth. It’s sentient indeed!

\section{Example: MNIST}

Leveraging a small MLP and VAE trained on MNIST data, this script trains a tokenizer that encodes geographical meaning in the latent embedding of VAE. Inspired by this tweet: https://x.com/savvyRL/status/1709698089500680264?s=20

https://gist.github.com/pat-alt/1c3d0d159b2cfa2a91a8d0c47ea3d28a

This tweet inspired me. Just as I had feared, even the small VAE I trained on MNIST a while ago is basically sentient. It is so over people!

\section{Ok, but truly what’s the point?}

The finding is not surprising but it is still interesting. In the context of mechanistic interpretability, it demonstrates that the black-box model has evidently learned plausible explanations for the data. Beyond that, in this particular example, the patterns in the latent space that we have just uncovered might actually be useful for downstream tasks. An interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model. To demonstrate this, let us consider a simple linear regression model for GDP growth. We will compare the performance of the following models: (1) regressing growth 
 on the latent embedding 
, (2) regression growth on the best subset of latent embeddings 
, (3) regressing growth on lagged growth, (4) regressing growth on lagged growth and the observed yield curve factors 
, (5) regressing growth on the best subset of yield curve factors 
, and, finally, (5) regressing growth on lagged growth and the best subset of latent embeddings 
.

@@

\section{LLM}

These results from the linear probe shown in @fig-fomc are certainly not unimpressive: even though *FOMC-RoBERTa* was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation. It is worth pointing out here that this model is substantially smaller than the models tested in @gurnee2023language. This begs the following question:

> Have we uncovered further evidence that LLMs "aren't mere stochastic parrots"? Has *FOMC-RoBERTa* developed an intrinsic understanding of the economy just by 'reading' central bank communications?

We find this very difficult to believe and we will now produce a counter-example to demonstrate that, if anything, these findings are very much in line with the parrot metaphor. If the results from the linear probe truly were indicative of some intrinsic understanding of the economy, then the probe should not be sensitive to random sentences that are most definitely not related to consumer prices.

To test this, we select the best-performing probe trained on the final-layer activations to predict changes in the CPI. We then make up sentences that fall into one of these four categories: *Inflation/Prices* (IP)---sentences about price inflation, *Deflation/Prices* (DP)---sentences about price deflation, *Inflation/Birds* (IB)---sentences about an inflation in the number of birds and *Deflation/Birds* (DB)---sentences about a deflation in the number of birds. A sensible sentence for category DP, for example, could be: "It is essential to bring inflation back to target to avoid drifting into deflation territory.". Analogically, we could construct the following sentence for the DB category: "It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.".

In light of the encouraging results for the probe in @fig-fomc, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category than for sentences in the DP category. If this is indicative of true intrinsic understanding, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing. More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict when using actual white noise as inputs.

To get to this last point, we also generate many probe predictions for samples of noise. Let $f: \mathcal{A}^k \mapsto \mathcal{Y}$ denote the linear probe that maps from the $k$-dimensional space spanned by $k$ first principal components of the final-layer activations to the output variable of interest (CPI growth). Then we sample $\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})$ for $i \in [1,1000]$ and compute the sample average. We repeat this process $10000$ times and compute the median-of-means to get an estimate for $\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]$, that is the predicted value of the probe conditional on white noise.

We propose the following hypothesis test to assess if the probe results provide evidence for an actual understanding of key economic relationships learned purely from text:

- *H0* (Null): The probe never predicts values that are statistically significantly different from $\mathbb{E}[f(\varepsilon)]$
- *H1* (Stochastic Parrots): The probe predicts values that are statistically significantly different from $\mathbb{E}[f(\varepsilon)]$ for sentences in all categories (IP,DP,IB,DB).
- *H2* (More than Mere Stochastic Parrots): The probe predicts values that are statistically significantly different from $\mathbb{E}[f(\varepsilon)]$ for sentences in categories IP and DP, but not for sentences in IB and DB.

>>>>

Hypotheses in computer science fields related to AI are often implicit. Often framed as informally that a given system is more accurate or efficient compared to other systems, and not typically articulated as they are in other fields, with specific conditions and in some instances numerical values or effect sizes assigned to each competing hypothesis. However with regards to claims that outputs of a system hint towards emerging qualities like AGI, hypotheses necessarily become more explicit: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).

Confirmation bias in hypothesis testing may manifest as a number of behaviors, as \cite{nickerson1998confirmation} reviews. Scientists may pay little to no attention to competing hypotheses or explanations, e.g. only considering the likelihood that outputs of system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false. Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a 'weak' or 'non-risky' experiment, a problem articulated in response to a number of scientific crises \cite{claesen2022severity}. If the threshold to accept a favored hypothesis is too low, observations consistent with the hypothesis are almost guaranteed, and in turn fail to severely test the claim in question. Thus, one is far more likely to show evidence in favor of their beliefs by posing a weak null hypotheses. Related to the formulation of hypotheses is the interpretation of evidence in favor of competing hypotheses, wherein people will interpret identical evidence differently based on their beliefs. As \cite{nickerson1998confirmation} reviews, individuals may place greater emphasis or milder criticism on evidence in support of their hypothesis, and lesser emphasis and greater criticism on evidence that opposes it.

@@

\subsection{Patterns in Latent Spaces and How to Find
Them}\label{patterns-in-latent-spaces-and-how-to-find-them}

To support my claim that observing patterns in latent spaces should not
generally surprise us, we will now go through a couple of simple
examples. To illustrate further that this phenomenon is neither
surprising nor unique to the field of computer science, I will draw on
my background in economics and finance in this section. We will start
with very simple examples to demonstrate that even small and simple
models can learn meaningful representations of the data. The final
example in \textbf{?@sec-ex-llm} is a bit more involved and closer in
spirit to the experiments conducted by Gurnee and Tegmark (2023). As we
go along, we will try to discuss both the benefits and potential
pitfalls of finding patterns in latent spaces.

\subsubsection{Example: Principal Component
Analysis}\label{example-principal-component-analysis}

In response to the claims made by Tegmark, numerous commentators on
social media have pointed out that even the simplest of models can
exhibit structure in their latent spaces. One of the most popular and
illustrative examples I remember from my time at the Bank of England is
yield curve decomposition through PCA. The yield curve is a popular tool
for investors and economists to gauge the health of the economy. It
plots the yields of bonds against their maturities. The slope of the
yield curve is often used as a predictor of future economic activity: a
steep yield curve is associated with a growing economy, while a flat or
inverted yield curve is associated with a contracting economy.

% To understand this better, let us go on a quick detour into economics
% and look at actual yield curves observed in the US during the Global
% Financial Crisis (GFC). Figure~\ref{fig-gfc-1} shows the yield curve of
% US Treasury bonds on 27 February 2007, which according to
% \href{https://money.cnn.com/2007/02/27/markets/markets_0630/index.htm?cnn=yes}{CNN}
% was a ``brutal day on Wall Street''.\footnote{The
%   \href{https://home.treasury.gov/resource-center/data-chart-center/interest-rates/TextView?type=daily_treasury_yield_curve&field_tdr_date_value_month=202310}{data}
%   is taken from the US Department of the Treasury.} This followed
% \href{https://money.cnn.com/2007/02/26/news/economy/greenspan/index.htm?postversion=2007022609}{reports}
% on the previous day of former Federal Reserve Chairman Alan Greenspan's
% warning that the US economy was at risk of a recession. The yield curve
% was inverted with a sharp negative spread between the 10-year and
% 3-month yields, indicative of the market's expectation of a recession.

% Figure~\ref{fig-gfc-2} shows the corresponding yield curve during the
% aftermath of the GFC on 20 April 2009. On that day the influential Time
% Magazine
% \href{https://content.time.com/time/business/article/0,8599,1890560,00.html}{reported}
% that the ``Banking Crisis is Over''. The yield curve was steeply sloped
% with a positive spread between the 10-year and 3-month yields,
% indicative of the market's expectation of a recovery. The overall level
% of the yield curve was still very low though, indicative of the fact
% that US economy had not fully recovered at that point.

@@

To compute the principal components we can decompose the matrix of
yields \(\mathbf{Y}\) into a product of its singular vectors and values:
\(\mathbf{Y}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\). I will not go into
the details here, because Professor Gilbert Strang has already done a
much better job than I ever could in his
\href{https://www.youtube.com/watch?v=mBcLRGuAFUk}{Linear Algebra
lectures}. To put this into the broader context of the article, however,
let us simply refer to \(\mathbf{U}\), \(\Sigma\) and
\(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve (they
are latent because they are not directly observable).

The top panel in Figure~\ref{fig-pca} shows the first two principal
components of the yield curves of US Treasury bonds over time. Vertical
stalks indicate the key dates during the onset and aftermath of the
crisis, which we discussed above. For both components, we can observe
some marked shifts between the two dates - but can we attribute any
meaning to these shifts? It turns out we can: for comparison, the bottom
panel in Figure~\ref{fig-pca} shows the average level and spread of the
yield curves over time. The first principal component is strongly
correlated with the level of the yield curve, while the second principal
component is strongly correlated with the spread of the yield curve. To
put it in AI-lingo:

\begin{quote}
The estimated latent embeddings of the yield curve are characterized by
patterns observed in the data.
\end{quote}

\begin{figure*}

\centering{

\includegraphics[width=0.8\textwidth]{results/figures/pca.png}

}

\caption{\label{fig-pca}Comparison of latent embeddings and observed
data of the US Treasury yield curve.}

\end{figure*}%

Not convinced? Let us use
\(\mathbf{Y}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\) in true autoencoder
fashion to reconstruct yield curves from principal components. Let
\(z_1\) denote the first principal component and consider the following:
we keep all other \(M-1\) principal components fixed at zero where \(M\)
denotes the total number of maturities; next we traverse the latent
space by varying the value of \(z_1\) over a fixed grid of length \(K\)
each time storing the full vector \(\mathbf{z}\); finally, we vertically
concatenate the vectors and end up with a matrix \(\mathbf{Z}\) of
dimension \((K \times M)\). To reconstruct yields, we simply multiply
\(Z\) by the singular values and right singular vectors:
\(\mathbf{Y}=\mathbf{Z}\Sigma\mathbf{V}^{\prime}\).

\subsubsection{Example: Deep Learning}\label{example-deep-learning}

@@

\paragraph{Data}\label{data}

To estimate economic growth we will rely on a quarterly
\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real
gross domestic product (GDP) provided by the Federal Reserve Bank of
St.~Louis. The data arrives in terms of levels of real GDP. In order to
estimate growth, we will transform the data into log differences. Since
our yield curve data is daily, we will need to aggregate it to the
quarterly frequency. To do this, we will simply take the average of the
daily yields for each maturity. We will also standardize yields since
deep learning models tend to perform better with standardized data.
Since COVID-19 was a huge structural break, we will also filter out all
observations after 2018.

\paragraph{Model}\label{model}

Let \(G_t\) denote growth and \(\mathbf{Y}_t\) denote the yield curve at
time \(t\). Then we are interested in a model for \(G_t\) conditional on
\(\mathbf{Y}_t\). Let \(\theta\) denote our model parameters then
formally we are interested in maximizing the likelihood
\(p_{\theta}(G_t|\mathbf{Y}_t)\).

To do this, we will use a simple
autoencoder architecture: the encoder will consist of a single fully
connected hidden layer with 32 neurons and a hyperbolic tangent
activation function. The bottleneck layer connecting the encoder to the
decoder is a fully connected layer with 6 neurons. This is the
compressed latent representation of the yield curve mentioned above. The
decoder will consist of two fully connected layers each with a
hyperbolic tangent activation function: the first layer will consist of
32 neurons and the second layer will have the same dimension as the
input data. The output layer will consist of a single neuron with a
linear activation function. The model will be trained using the mean
squared error loss function and the Adam optimizer.

\paragraph{Linear Probe}\label{linear-probe}

The results are shown in Figure~\ref{fig-dl-results}. The top panel
shows the actual GDP growth and fitted values from the autoencoder
model. We observe that the model captures the relationship between
economic growth and the yield curve reasonably well. As discussed above,
we also know that the relationship between economic growth and the yield
curve is characterized by two main factors: the level and the spread.
Since the model itself is fully characterized by its parameters, we
would expect that these two important factors are reflected somewhere in
the latent parameter space.

The bottleneck layer seems like a good place to start looking. To get
the latent embeddings \(\mathbf{A}_t\) at time \(t\), we simply pass the
yield curve data \(\mathbf{Y}_t\) through the encoder. Next, we follow
Gurnee and Tegmark (2023) and use a linear probe to regress the observed
yield curve factors on the latent embeddings. Let \(F_t\) denote the
vector containing the two factors of interest in time \(t\):
\(f_t^{\text{spread}}\) and \(f_t^{\text{level}}\). Formally, we are
interested in the following regression model:
\(p_{w}(F_t|\mathbf{A}_t)\) where \(w\) denotes the regression
parameters. Following Gurnee and Tegmark (2023), we use Ridge regression
with \(\lambda\) set to \(0.1\). Using the estimated regression
parameters \(\hat{w}\) we can then predict the yield curve factors from
the latent embeddings: \(\hat{F}_t=\hat{w}^{\prime}\mathbf{A}_t\).

The results of this experiment are shown in the bottom panel of
Figure~\ref{fig-dl-results}. Solid lines show the observed yield curve
factors over time, while dashed lines show predicted values. We find
that the latent embeddings predict the two yield curve factors
reasonably well, in particular the spread. To put this in true
hype-lingo:

\begin{quote}
Not just a parrot! Our autoencoder neural network has an implicit
understanding of the yield curve factors and their relationship with
economic growth. It's sentient indeed!
\end{quote}

\begin{figure}

\centering{

\includegraphics[width=0.5\textwidth]{results/figures/dl.png}

}

\caption{\label{fig-dl-results}The top panel shows the actual GDP growth
and fitted values from the autoencoder model. The bottom panel shows the
observed average level and spread of the yield curve (solid) along with
the predicted values from the linear probe based on the latent embedding
(dashed).}

\end{figure}%

\paragraph{Ok, but truly what's the
point?}\label{ok-but-truly-whats-the-point}

The finding is not surprising but it is still interesting. In the
context of mechanistic interpretability, it demonstrates that the
black-box model has evidently learned plausible explanations for the
data. Beyond that, in this particular example, the patterns in the
latent space that we have just uncovered might actually be useful for
downstream tasks. An interesting idea could be to use the latent
embeddings as features in a more traditional and interpretable
econometric model. To demonstrate this, let us consider a simple linear
regression model for GDP growth. We will compare the performance of the
following models: (1) regressing growth \(G_t\) on the latent embedding
\(A_t\), (2) regression growth on the best subset of latent embeddings
\(A_t\), (3) regressing growth on lagged growth, (4) regressing growth
on lagged growth and the observed yield curve factors \(F_t\), (5)
regressing growth on the best subset of yield curve factors \(F_t\),
and, finally, (5) regressing growth on lagged growth and the best subset
of latent embeddings \(A_t\).

@@

\input{results/regression.tex}

\subsubsection{Example: Large Language Model}\label{ex-llm}

To round up this section, we will jump back on the hype train and
consider an example involving an LLM. In particular, we will closely
follow the approach in Gurnee and Tegmark (2023) and apply it to a novel
financial dataset: the \emph{Trillion Dollar Words} dataset introduced
by Shah, Paturi, and Chava (2023). The dataset contains a curated
selection of sentences formulated by central bankers of the US Federal
Reserve and communicated to the public in speeches, meeting minutes and
press conferences. The authors of the paper use this dataset to train
LLMs to classify sentences as either `dovish', `hawkish' or `neutral'.
To this end, they first manually annotate a subsample of the available
data and then fine-tune various foundation models. Their model of
choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa (Liu et al.
2019)), achieves an \(F_1\) score of around \(>0.7\) for the
classification task. To illustrate the potential usefulness of the
learned classifier, they use predicted labels for the entire dataset to
compute an ad-hoc, count-based measure of `hawkishness'. They then go on
to show that this measure correlates with key economic indicators in the
expected direction: when inflationary pressures rise, the measured level
of `hawkishness' increases as central bankers need to raise interest
rates to bring inflation back to target.

\paragraph{Linear Probes}\label{linear-probes}

Instead of computing a measure based on predicted labels, we can use
linear probes to assess if the fine-tuned model has learned associative
patterns between central bank communications and key economic
indicators. To this end, I have further pre-processed the data provided
by Shah, Paturi, and Chava (2023) and used their proposed model to
compute layer-wise embeddings for all available sentences. I have made
these available and easily accessible through a small Julia package:
\href{https://github.com/pat-alt/TrillionDollarWords.jl}{TrillionDollarWords.jl}.
For each layer, I have then computed linear probes on two inflation
indicators---the Consumer Price Index (CPI) and the Producer Price Index
(PPI)---as well as US Treasury yields at different levels of maturity.
To mitigate issues related to over-parameterization, I follow the
recommendation in Alain and Bengio (2018) to first reduce the
dimensionality of the embeddings each time. In particular, linear probes
are restricted to the first 128 principal components of the embeddings
of each layer.

Detailed results of this experiment will be released in an upcoming
paper. Figure~\ref{fig-fomc} highlights the result for the linear probe
on the CPI. The chart shows various performance measures plotted against
\emph{FOMC-RoBERTa}'s \(n\)-th layer. Shaded areas show the variation
across cross-validation folds, where we have used an expanding window
approach to split the time series. To avoid look-ahead bias, we run PCA
separately for each training set. Results for the linear probes are
shown along with results for autoregressive models (AR) used as our
baseline.

The first column shows the correlations between model predictions and
observed inflation. One can observe that this correlation is strictly
positive for the baseline models and the linear probes. Consistent with
the findings in Gurnee and Tegmark (2023) and Alain and Bengio (2018),
we also observe that this correlation tends to be higher for layers near
the end of the transformer model. Those layers produce predictions that
are more positively correlated with observed inflation than those
produced by the corresponding autoregressive models.

Similarly, we observe that the (root) mean squared error of the linear
probe is largely on par with the baseline. The average prediction error
is gradually reduced as we move along the horizontal axis, which is
again indicative of the notion that layers closer to the end of the
neural network have distilled more useful representations.

\begin{figure}

\centering{

\includegraphics[width=0.5\textwidth]{results/figures/measures_CPI (n_pc=128).png}

}

\caption{\label{fig-fomc}Various performance measures (correlation and
(root) mean squared error) plotted against \emph{FOMC-RoBERTa}'s
\(n\)-th layer. Shaded areas indicate variation across cross-validation
folds, where we have used an expanding window approach to split the time
series. Results for the linear probes are shown along with results for
autoregressive models (AR). The optimal lag length was determined using
the Bayes Information Criterium.}

\end{figure}%

\paragraph{Stochastic Parrots After
All?}\label{stochastic-parrots-after-all}

These results from the linear probe shown in Figure~\ref{fig-fomc} are
certainly not unimpressive: even though \emph{FOMC-RoBERTa} was not
explicitly trained to uncover associations between central bank
communications and the level of consumer prices, it appears that the
model has distilled representations that can be used to predict
inflation. It is worth pointing out here that this model is
substantially smaller than the models tested in Gurnee and Tegmark
(2023). This begs the following question:

\begin{quote}
Have we uncovered further evidence that LLMs ``aren't mere stochastic
parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic understanding
of the economy just by `reading' central bank communications?
\end{quote}

Personally, I am having a very hard time believing this. To argue my
case, I will now produce a counter-example demonstrating that, if
anything, these findings are very much in line with the parrot metaphor.
The counter-example is based on the following premise: if the results
from the linear probe truly were indicative of some intrinsic
understanding of the economy, then the probe should not be sensitive to
random sentences that are most definitely not related to consumer
prices.

To test this, I select the best-performing probe trained on the
final-layer activations to predict changes in the CPI. I then make up
sentences that fall into one of these four categories:
\emph{Inflation/Prices} (IP)---sentences about price inflation,
\emph{Deflation/Prices} (DP)---sentences about price deflation,
\emph{Inflation/Birds} (IB)---sentences about an inflation in the number
of birds and \emph{Deflation/Birds} (DB)---sentences about a deflation
in the number of birds. A sensible sentence for category DP, for
example, could be: ``It is essential to bring inflation back to target
to avoid drifting into deflation territory.''. Analogically, we could
construct the following sentence for the DB category: ``It is essential
to bring the numbers of doves back to target to avoid drifting into
dovelation territory.''.

In light of the encouraging results for the probe in
Figure~\ref{fig-fomc}, we should expect the probe to predict higher
levels of inflation for activations for sentences in the IP category
than for sentences in the DP category. If this was indicative of true
intrinsic understanding, we would not expect to see any significant
difference in predicted inflation levels for sentences about birds,
independent of whether or not their numbers are increasing. More
specifically, we would not expect the probe to predict values for
sentences about birds that are substantially different from the values
it can be expected to predict when using actual white noise as inputs.

To get to this last point, I also generate many probe predictions for
samples of noise. Let \(f: \mathcal{A}^k \mapsto \mathcal{Y}\) denote
the linear probe that maps from the \(k\)-dimensional space spanned by
\(k\) first principal components of the final-layer activations to the
output variable of interest (CPI growth in this case). Then I sample
\(\varepsilon_i \sim \mathcal{N}(\mathbf{0},\mathbf{I}^{(k \times k)})\)
for \(i \in [1,1000]\) and compute the sample average. I repeat this
process \(10000\) times and compute the median-of-means to get an
estimate for \(\mathbb{E}[f(\varepsilon)]=\mathbb{E}[y|\varepsilon]\),
that is the predicted value of the probe conditional on white noise.

Next, I propose the following hypothesis test as a minimum viable
testing framework to assess if the probe results (may) provide evidence
for an actual understanding of key economic relationships learned purely
from text:

\begin{proposition}[Parrot
Test]\protect\hypertarget{prp-line}{}\label{prp-line}

~

\begin{itemize}
\tightlist
\item
  \emph{H0} (Null \emph{WEAK}): The probe never predicts values that are
  statistically significantly different from
  \(\mathbb{E}[f(\varepsilon)]\)
\item
  \emph{H1} (Stochastic Parrots \emph{RISKIER}): The probe predicts
  values that are statistically significantly different from
  \(\mathbb{E}[f(\varepsilon)]\) for sentences in all categories
  (IP,DP,IB,DB).
\item
  \emph{H2} (More than Mere Stochastic Parrots \emph{RISKIEST}): The
  probe predicts values that are statistically significantly different
  from \(\mathbb{E}[f(\varepsilon)]\) for sentences in categories IP and
  DP, but not for sentences in IB and DB.
\end{itemize}

\end{proposition}

To be clear, if in such a test we did find substantial evidence in
favour of rejecting both \emph{HO} and \emph{H1}, this would not
automatically imply that \emph{H2} is true. But to even continue
investigating if based on having learned meaningful representation the
underlying LLM is more than just a parrot, it should be able to pass
this simple test.

In this particular case, Figure~\ref{fig-attack} demonstrates that we
find some evidence to reject \emph{H0} but not \emph{H1} for
\emph{FOMC-RoBERTa}. The median linear probe predictions for sentences
about inflation and deflation are indeed substantially higher and lower,
respectively than for random noise. Unfortunately, the same is true for
sentences about the inflation and deflation in the number of birds,
albeit to a somewhat lower degree.

I should note that the number of sentences in each category is very
small here (10), so the results in Figure~\ref{fig-attack} cannot be
used to establish statistical significance. That being said, even a
handful of convincing counter-examples should be enough for us to
seriously question the claim that results from linear probes provide
evidence against the parrot metaphor. In fact, if there's even a single
sentence for which any

\begin{quote}
Strong alternative hypothesis to address own confirmation bias
Articulating a test to better differentiate
\end{quote}

\begin{figure}

\centering{

\includegraphics[width=0.5\textwidth]{results/figures/attack_inflation.png}

}

\caption{\label{fig-attack}Probe predictions for sentences about
inflation of prices (IP), deflation of prices (DP), inflation of birds
(IB) and deflation of birds (DB). The vertical axis shows predicted
inflation levels subtracted by the average predicted inflation level for
the whole sample period (train and test set).}

\end{figure}%
